Prefetching SPARQL query cacher

### 2016-07-28 dissertation/cache-proxy.tex
large cleanup of join rotation

\subsection{Understanding the Query Planning}

@@

Merging the quad plans into a single SPARQL BGP plan is not
straightforward in the presence of a quad pattern that is to be looked
up in e.g. a cache. If we assign the letters $A$, $B$, $C$ and $D$ to
quad patterns \todo{correctly expressed?} in the order in
Example~\ref{ex:bgp4}, the initial join tree produced by parsing may
be $ ((A \bowtie B )\bowtie C) \bowtie D) $. Now, we are interested in
a plan on the form $A \bowtie BCD$ (all commutations will be
generated). To facilitate this, we employ \emph{join rotation}. A
simple example of join rotation would be this: We first generate
subplans for both $((A \bowtie B )\bowtie C) \bowtie D) $ and $ (A
\bowtie B )\bowtie (C \bowtie D) $. Then, $C$ and $D$ can be merged to
a SPARQL BGP to $ (A \bowtie B )\bowtie CD $. Another rotation can
then be performed: $ A \bowtie (B \bowtie CD) $ and then coalesced to
the final result. The planner will generate alternative plans for both
rotated and unrotated plans, and cost is estimated for both, so we
have to be careful to design the cost model to allow the rotation to
result in SPARQL BGP plans.

@@

It is interesting to detail the function of
\pmodule{AtteanX::API::JoinRotatingPlanner}. Its main function is to
produce alternatives for join query plans. For example, if we have
three quad patterns $A$, $B$ and $C$, plans for both $ ( A \bowtie B)
\bowtie C $ and $ A \bowtie ( B \bowtie C ) $ will be generated. In
some cases, it may be possible to reduce a join to a more fundamental
plan type, and this may be done next. In our case, this is done in
\pmodule{AtteanX::QueryPlanner::Cache}, where adjacent quad patterns
are coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
coalesced into larger BGP plans, so that they can be evaluated by the
remote SPARQL endpoint as a single unit. This will minimise the number
of HTTP requests, in much the same way as the exclusive groups of
\cite{springerlink:10.1007/978-3-642-25073-6-38}. Note, however, that
the generated plans are subject to cost estimation, so whether a large
BGP plan will be evaluated remotely, or as a join of local cached
results and a smaller remote BGP, or in the case of
\pmodule{AtteanX::QueryPlanner::Cache::LDF}, a remote Triple Pattern
Fragments server, is up to the cost estimation. This will be discussed
in detail in Section~\ref{sec:costheuristics}.

>>>>

\subsection{Understanding the Query Planning}\label{sec:understandquery}

@@

Merging the quad plans into a single SPARQL BGP plan is not
straightforward in the presence of a quad pattern that is to be looked
up in e.g. a cache. If we assign the letters $A$, $B$, $C$ and $D$ to
quad patterns \todo{correctly expressed?} in the order in
Example~\ref{ex:bgp4}, the initial join tree produced by parsing may
be $ ((A \bowtie B )\bowtie C) \bowtie D) $. Now, we are interested in
a plan on the form $A \bowtie BCD$, i.e. $BCD$ is reduced to a more
fundamental plan type, in this example a SPARQL BGP. This will
minimise the number of HTTP requests, in much the same way as the
exclusive groups of
\cite{springerlink:10.1007/978-3-642-25073-6-38}. To facilitate this,
we employ \emph{join rotation}. A simple example of join rotation
would be this: We first generate subplans for both $((A \bowtie B
)\bowtie C) \bowtie D) $ and $ (A \bowtie B )\bowtie (C \bowtie D)
$. Then, $C$ and $D$ can be merged to a SPARQL BGP to $ (A \bowtie B
)\bowtie CD $. Another rotation can then be performed: $ A \bowtie (B
\bowtie CD) $ and then coalesced to the final result. The planner will
generate alternative plans for both rotated and unrotated plans, and
all generated plans are subject to cost estimation, so whether a large
BGP plan will be evaluated remotely, or as a join of local cached
results and a smaller remote BGP, or a remote Triple Pattern Fragments
server, is up to the cost estimation. This will be discussed in detail
in Section~\ref{sec:costheuristics}.

@@

The implementation of the join rotation detailed in
Section~\ref{sec:understandquery} is in the role
\pmodule{AtteanX::API::JoinRotatingPlanner}. In the planner
\pmodule{AtteanX::QueryPlanner::Cache} adjacent quad patterns are
coalesced into Basic Graph Pattern (BGP) plans, and BGP plans are
coalesced into larger BGP plans, so that they can be evaluated by the
remote SPARQL endpoint as a single unit. The extension
\pmodule{AtteanX::QueryPlanner::Cache::LDF} adds the option of
querying remote LDF servers.
